---
title: Assignment 4
author: Matthew Kielar
date: "`r Sys.Date()`"
output:
    bookdown::pdf_document2:
        number_sections: false
        toc: false
---

```{r, echo = TRUE, message = FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
```

```{r, include = FALSE, echo = FALSE}
plot_theme <- theme(
    plot.title = element_text(
        size = 16, face = "bold", hjust = 0,
        margin = margin(b = 0.1, unit = "cm")),
    plot.background = element_rect(
        fill = "#FFFFFF"),
    axis.title = element_text(
        size = 12, face = "bold"),
    axis.text = element_text(
        size = 12),
    axis.ticks.length = unit(0.25, units = "cm"),
    axis.line = element_line(
        linewidth = 0.5, color = "#000000"),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(
        linewidth = 0.1, color = "gray50"),
    panel.grid.major.y = element_line(
        linewidth = 0.1, color = "gray50"),
    panel.background = element_rect(
        fill = "#FFFFFF"),
    panel.border = element_rect(
        fill = NA, color = "#000000", linewidth = 2.0)
)

theme_set(plot_theme)
```

## Question 2

__Part a)__

$$
\begin{aligned}
\gamma(0) &= \int_{0}^{\pi} \cos(0 \cdot \omega)f(\omega)d\omega \\[15pt]
&= \int_{0}^{\pi} f(\omega)d\omega \\[15pt]
&= \int_{0}^{\pi} \frac{1}{8\pi} \left(1 - \frac{32}{35}\cos(\omega) + \frac{3}{5}\cos(2\omega)\right)d\omega \\[15pt]
&= \frac{1}{8\pi} \int_{0}^{\pi} d\omega - \frac{32}{200\pi} \int_{0}^{\pi} \cos(\omega)d\omega + \frac{3}{40\pi} \int_{0}^{\pi} \cos(2\omega)d\omega \\
\end{aligned}
$$

The integrals can then be evaluated separately

$$
\begin{aligned}
\frac{1}{8\pi} \int_{0}^{\pi} d\omega &= \frac{1}{8\pi} \cdot \omega \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{1}{8\pi}(\pi - 0) \\[15pt]
&= \frac{1}{8}
\end{aligned}
$$

$$
\begin{aligned}
\frac{32}{200\pi} \int_{0}^{\pi} \cos(\omega) d\omega &= \frac{32}{200\pi} \cdot \sin(\omega) \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{32}{200\pi} \left(\sin(\pi) - \sin(0)\right) \\[15pt]
&= \frac{32}{200\pi} \cdot (0 - 0) \\[15pt]
&= 0
\end{aligned}
$$

$$
\begin{aligned}
\frac{3}{40\pi} \int_{0}^{\pi} \cos(2\omega)d\omega &= \frac{3}{40\pi} \cdot \left(\frac{1}{2}\sin(2\omega)\right) \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{3}{40\pi} \cdot \left(\frac{1}{2}\sin(2\pi) - \frac{1}{2}\sin(0)\right) \\[15pt]
&= \frac{3}{40\pi} \cdot \left(0 - 0\right) \\[15pt]
&= 0
\end{aligned}
$$

So we get 

$$\underbrace{\frac{1}{8\pi} \int_{0}^{\pi}d\omega}_\textrm{1/8} - \underbrace{\frac{32}{200\pi} \int_{0}^{\pi} \cos(\omega)d\omega}_\textrm{0} + \underbrace{\frac{3}{40\pi} \int_{0}^{\pi} \cos(2\omega)d\omega}_\textrm{0} = \frac{1}{8}$$

## Question 3

__Part a)__

```{r, echo = TRUE, message = FALSE}
pelts <- read_table("./data/pelt.txt", col_names = FALSE) %>%
    rename(
        sales = `X1`) %>%
    mutate(
        year = seq(year("1857-01-01"), year("1910-01-01"))) %>%
    relocate(year, .before = sales)

pelts_chart <- ggplot(
    data = pelts,
    aes(x = year, y = sales)
) +
    geom_line(
        linewidth = 1.2,
        color = "#FF8833"
) +
    labs(
        x = "Year",
        y = "Sales",
        title = "Pelt Sales Between 1857 and 1910"
)

print(pelts_chart)
```

This looks like a pretty standard time series with an obvious seasonal effect.
However, the seasonal variation occurs over periods of years, rather than months
like we've usually seen so far. Non-seasonal levels largely remain constant 
throughout the entire time period.

__Part b)__

```{r, echo = TRUE, message = FALSE}
pelts_spec <- spec.pgram(
    ts(pelts$sales, start = c(1857), frequency = 1),
    log = "no",
    plot = FALSE)

spec_data <- tibble(
    freq = seq(0, 0.5, length.out = length(pelts_spec$spec)),
    spec = pelts_spec$spec
)

spec_chart <- ggplot(
    data = spec_data,
    aes(x = freq, y = spec)
) +
    geom_line(
        linewidth = 1.0,
        color = "#FF0652"
) +
    labs(
        x = "Frequency",
        y = "Spectrum",
        title = "Spectral Density of Pelt Sales"
)

print(spec_chart)
```


__Part c)__

```{r, echo = TRUE, message = FALSE}
#' @param series A univariate time-series
#' @param p Placeholder constant

fourier_freq <- function(series, p) {
    # Input checks
    if (p > length(series) / 2) {
        stop("p cannot be greater than N/2")
    }
    if (is.ts(series) == FALSE) {
        warning("'series' is not of type 'ts', coercing to ts")
        series <- as.ts(series)
    }
    fft <- fft(series)
    return(head(fft))
}
```