---
title: Assignment 4
author: Matthew Kielar
date: "`r Sys.Date()`"
output:
    bookdown::pdf_document2:
        number_sections: false
        toc: false
        highlight: kate
---

```{r, echo = TRUE, message = FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
```

```{r, include = FALSE, echo = FALSE}
plot_theme <- theme(
    plot.title = element_text(
        size = 16, face = "bold", hjust = 0,
        margin = margin(b = 0.1, unit = "cm")),
    plot.background = element_rect(
        fill = "#FFFFFF"),
    axis.title = element_text(
        size = 12, face = "bold"),
    axis.text = element_text(
        size = 12),
    axis.ticks.length = unit(0.25, units = "cm"),
    axis.line = element_line(
        linewidth = 0.5, color = "#000000"),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(
        linewidth = 0.1, color = "gray50"),
    panel.grid.major.y = element_line(
        linewidth = 0.1, color = "gray50"),
    panel.background = element_rect(
        fill = "#FFFFFF"),
    panel.border = element_rect(
        fill = NA, color = "#000000", linewidth = 2.0),
    legend.title = element_text(
        size = 12, face = "bold"),
    legend.text = element_text(
        size = 12),
    legend.position = "top",
    legend.justification = "left"
)

theme_set(plot_theme)
```

## Question 1

__Part a)__ $\medskip$

For the above AR(2) process, we have $\phi(B) = (1 - 1.1B + 0.3B^2)$, so we get

$$
\begin{aligned}
f(\omega) &= \frac{\sigma_Z^2}{\pi} \cdot \frac{1}{\lvert\phi(e^{i\omega})\rvert^2} \\[15pt]
&= \frac{\sigma_Z^2}{\pi} \cdot \frac{1}{\lvert 1 - 1.1e^{i\omega} + 0.3e^{2i\omega}\rvert^2} \\[15pt]
&= \frac{\sigma_Z^2}{\pi} \cdot \frac{1}{\lvert 1 - 1.1(\cos\omega + i\sin\omega) + 0.3(\cos 2\omega + i\sin 2\omega)\rvert^2} \\[15pt]
&= \frac{\sigma_Z^2}{\pi} \cdot \frac{1}{\lvert 1 - 1.1\cos\omega - 1.1i\sin\omega + 0.3\cos 2\omega + 0.3i\sin 2\omega \rvert^2} \\[15pt]
&= \frac{\sigma_Z^2}{\pi} \cdot \frac{1}{(1 - 1.1\cos\omega + 0.3\cos 2\omega)^2 + (-1.1\sin\omega + 0.3\sin 2\omega)^2} \\[15pt]
&= \frac{\sigma_Z^2}{\pi[2.3 - 2.86\cos\omega + 0.6\cos 2\omega]} \\[15pt]
\end{aligned}
$$

Then we can calculate the variance of $X_t$ using a shortcut

$$\sigma_X^2 = \frac{\sigma_Z^2}{1 - \rho(1)\phi_1 - \rho(2)\phi_2} \\[15pt]$$

With the given acf, we have $\rho(1) = 0.846$ and $\rho(2) = 0.631$, so we get

$$
\begin{aligned}
\sigma_X^2 &= \frac{\sigma_Z^2}{1 - 0.846(1.1) - 0.631(-0.3)} \\[15pt]
&= \frac{\sigma_Z^2}{0.259} \\
\end{aligned}
$$

Plugging this into $f^{\ast}(\omega) = f(\omega)/\sigma_X^2$ gives

$$
\begin{aligned}
f^{\ast}(\omega) &= \frac{\sigma_Z^2}{\pi[2.3 - 2.86\cos\omega +0.6\cos 2\omega]} \cdot \frac{0.259}{\sigma_Z^2} \\[15pt]
&= \frac{0.259}{\pi[2.3 - 2.86\cos\omega + 0.6\cos 2\omega]} \\[15pt]
\end{aligned}
$$

Note: the derivation above used a method outlined on page 161 of the textbook. $\medskip$

__Part b)__ $\medskip$

As already derived in a), the power spectral density function of $\{X_t\}_{t \in \mathbb{Z}}$ is

$$f(\omega) = \frac{\sigma_Z^2}{\pi[2.3 - 2.86\cos\omega + 0.6\cos 2\omega]}$$

__Part c)__ 

```{r, echo = TRUE, message = FALSE}
norm_spec <- function(w) {
    spec <- 0.259 / (pi * (2.3 - 2.86 * cos(w) + 0.6 * cos(2 * w)))
    return(spec)
}

spec_data <- tibble(
    w = seq(1, pi, length.out = 100)) %>%
    mutate(
        spectrum = norm_spec(w))

norm_spec_chart  <- ggplot(
    data = spec_data,
    aes(x = w / (2 * pi), y = spectrum)
) +
    geom_line(
        linewidth = 1.2,
        color = "#FF0652"
) +
    labs(
        x = "Frequency",
        y = "Normalized Spectrum",
        title = "Normalized Spectral Density of an AR(2) Process"
)

print(norm_spec_chart)
```

## Question 2

__Part a)__

$$
\begin{aligned}
\gamma(0) &= \int_{0}^{\pi} \cos(0 \cdot \omega)f(\omega)d\omega \\[15pt]
&= \int_{0}^{\pi} f(\omega)d\omega \\[15pt]
&= \int_{0}^{\pi} \frac{1}{8\pi} \left(1 - \frac{32}{35}\cos(\omega) + \frac{3}{5}\cos(2\omega)\right)d\omega \\[15pt]
&= \frac{1}{8\pi} \int_{0}^{\pi} d\omega - \frac{32}{200\pi} \int_{0}^{\pi} \cos(\omega)d\omega + \frac{3}{40\pi} \int_{0}^{\pi} \cos(2\omega)d\omega \\[15pt]
\end{aligned}
$$

The integrals here can then be calculated separately

$$
\begin{aligned}
\frac{1}{8\pi} \int_{0}^{\pi} d\omega &= \frac{1}{8\pi} \cdot \omega \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{1}{8\pi}(\pi - 0) \\[15pt]
&= \frac{1}{8}
\end{aligned}
$$

$$
\begin{aligned}
\frac{32}{200\pi} \int_{0}^{\pi} \cos(\omega) d\omega &= \frac{32}{200\pi} \cdot \sin(\omega) \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{32}{200\pi} \left(\sin(\pi) - \sin(0)\right) \\[15pt]
&= \frac{32}{200\pi} \cdot (0 - 0) \\[15pt]
&= 0
\end{aligned}
$$

$$
\begin{aligned}
\frac{3}{40\pi} \int_{0}^{\pi} \cos(2\omega)d\omega &= \frac{3}{40\pi} \cdot \left(\frac{1}{2}\sin(2\omega)\right) \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{3}{40\pi} \cdot \left(\frac{1}{2}\sin(2\pi) - \frac{1}{2}\sin(0)\right) \\[15pt]
&= \frac{3}{40\pi} \cdot \left(0 - 0\right) \\[15pt]
&= 0
\end{aligned}
$$

So we get 

$$\underbrace{\frac{1}{8\pi} \int_{0}^{\pi}d\omega}_\textrm{1/8} - \underbrace{\frac{32}{200\pi} \int_{0}^{\pi} \cos(\omega)d\omega}_\textrm{0} + \underbrace{\frac{3}{40\pi} \int_{0}^{\pi} \cos(2\omega)d\omega}_\textrm{0} = \frac{1}{8}$$

__Part b)__

$$
\begin{aligned}
\gamma(1) &= \int_{0}^{\pi} \cos(\omega)f(\omega)d\omega \\[15pt]
&= \frac{1}{8\pi} \int_{0}^{\pi} \left(\cos(\omega) - \frac{32}{25}\cos^2(\omega) + \frac{3}{5}\cos(\omega)\cos(2\omega)\right) \\[15pt]
&= \frac{1}{8\pi} \int_{0}^{\pi} \cos(\omega)d\omega - \frac{32}{200\pi} \int_{0}^{\pi} \cos^2(\omega)d\omega + \frac{3}{40\pi} \int_{0}^{\pi} \cos(\omega)\cos(2\omega)d\omega \\[15pt]
\end{aligned}
$$

Like before, it's probably a better idea to calculate the integrals individually

$$
\begin{aligned}
\frac{1}{8\pi} \int_{0}^{\pi} \cos(\omega)d\omega &= \frac{1}{8\pi} \cdot (\sin(\omega)) \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{1}{8\pi} \cdot (\sin(\pi) - \sin(0)) \\[15pt]
&= 0
\end{aligned}
$$

$$
\begin{aligned}
\frac{32}{200\pi} \int_{0}^{\pi} \cos^2(\omega) &= \frac{32}{200\pi} \int_{0}^{\pi} \left(\frac{1}{2} \cos(2\omega) + \frac{1}{2} \right)d\omega \\[15pt]
&= \frac{32}{200\pi} \cdot \left(\frac{1}{4} \sin(2\omega) + \frac{1}{2}\omega \right) \bigg\rvert_{0}^{\pi} \\[15pt]
&=\frac{32}{200\pi} \cdot \left[\left(\frac{1}{4} \sin(2\pi) + \frac{\pi}{2}\right) - \left(\frac{1}{4} \sin(0) + 0 \right)\right] \\[15pt]
&= \frac{32}{200\pi} \cdot \frac{\pi}{2} \\[15pt]
&= \frac{2}{25}
\end{aligned}
$$

$$
\begin{aligned}
\frac{3}{40\pi} \int_{0}^{\pi} \cos(\omega)\cos(2\omega)d\omega &= \frac{3}{40\pi} \int_{0}^{\pi} \frac{1}{2}(\cos(3\omega) + \cos(\omega))d\omega \\[15pt]
&= \frac{3}{40\pi} \cdot \left(\frac{1}{6}\sin(3\omega) + \frac{1}{2} \sin(\omega)\right) \bigg\rvert_{0}^{\pi} \\[15pt]
&= \frac{3}{40\pi} \cdot \left[\left(\frac{1}{6}\sin(3\pi) + \frac{1}{2}\sin(\pi) \right) - \left(\frac{1}{6}\sin(0) + \frac{1}{2}\sin(0)\right) \right] \\[15pt]
&= \frac{3}{40\pi} \cdot (0 - 0) \\[15pt]
&= 0
\end{aligned}
$$

So we get

$$\underbrace{\frac{1}{8\pi} \int_{0}^{\pi} \cos(\omega)d\omega}_\textrm{0} - \underbrace{\frac{32}{200\pi} \int_{0}^{\pi} \cos^2(\omega)d\omega}_\textrm{2/25} + \underbrace{\frac{3}{40\pi} \int_{0}^{\pi} \cos(\omega)\cos(2\omega)d\omega}_\textrm{0} = -\frac{2}{25}$$

\clearpage

__Part c)__

$$
\begin{aligned}
\gamma(2) &= \int_0^\pi \cos(2\omega)f(\omega)d\omega \\[15pt]
&= \frac{1}{8\pi} \int_0^\pi \left(\cos(2\omega)d\omega - \frac{32}{25}\int_0^\pi \cos(2\omega)\cos(\omega)d\omega + \frac{3}{5}\int_0^\pi \cos^2(2\omega)d\omega\right) \\[15pt]
&= \frac{1}{8\pi} \int_0^\pi \cos(2\omega)d\omega - \frac{32}{200\pi}\int_0^\pi \cos(2\omega)\cos(\omega)d\omega + \frac{3}{40\pi} \int_0^\pi \cos^2(2\omega)d\omega \\[15pt]
\end{aligned}
$$

Calculating the integrals separately means we get

$$
\begin{aligned}
\frac{1}{8\pi} \int_0^\pi \cos(2\omega)d\omega &= \frac{1}{8\pi} \cdot \left(\frac{1}{2} \sin(2\omega) \right) \bigg\rvert_0^\pi \\[15pt]
&= \frac{1}{8\pi} \cdot \left(\frac{1}{2}\sin(2\pi) - \frac{1}{2}\sin(0) \right) \\[15pt]
&= \frac{1}{8\pi} \cdot (0 - 0 ) \\[15pt]
&= 0
\end{aligned}
$$

$$
\begin{aligned}
\frac{32}{200\pi} \int_0^\pi \cos(2\omega)\cos(\omega)d\omega &= \int_0^\pi \frac{1}{2}(\cos(3\omega) + \cos(\omega))d\omega \\[15pt]
&= \frac{32}{200\pi} \cdot \left(\frac{1}{6}\sin(3\omega) + \frac{1}{2}\sin(\omega)\right) \bigg\rvert_0^\pi \\[15pt]
&= \frac{32}{200\pi} \cdot \left[\left(\frac{1}{6}\sin(3\pi) + \frac{1}{2}\sin(\pi)\right) - \left(\frac{1}{6}\sin(0) + \frac{1}{2}\sin(0)\right)\right] \\[15pt]
&= \frac{32}{200\pi} \cdot (0 - 0) \\[15pt]
&= 0
\end{aligned}
$$

$$
\begin{aligned}
\frac{3}{40\pi} \int_0^\pi \cos^2(2\omega)d\omega &= \frac{3}{40\pi} \int_0^\pi \frac{1}{2}(\cos(4\omega) + 1)d\omega \\[15pt]
&= \frac{3}{40\pi} \cdot \left(\frac{1}{8}\sin(4\omega) + \frac{1}{2}\omega \right) \bigg\rvert_0^\pi \\[15pt]
&= \frac{3}{40\pi} \cdot \left[\left(\frac{1}{8}\sin(4\pi) + \frac{\pi}{2} \right) - \left(\frac{1}{8}\sin(0) + 0 \right)\right] \\[15pt]
&= \frac{3}{40\pi} \cdot \frac{\pi}{2} \\[15pt]
&= \frac{3}{80}
\end{aligned}
$$

So we get

$$\gamma(2) = \underbrace{\frac{1}{8\pi} \int_0^\pi \cos(2\omega)d\omega}_\textrm{0} - \underbrace{\frac{32}{200\pi} \int_0^\pi \cos(2\omega)\cos(\omega)d\omega}_\textrm{0} + \underbrace{\frac{3}{40\pi} \int_0^\pi \cos^2(2\omega)d\omega}_\textrm{3/80} = \frac{3}{80}$$

__Part d)__ $\medskip$

The function below computes the definite integral of the cosine transform of $f(\omega)$ for 
a given lag $k$ and returns the result.

```{r, echo = TRUE, message = FALSE}
acvf <- function(k) {
    gamma_k <- function(w) {
        (1/(8*pi)) * (cos(k*w) - 1.28*cos(k*w)*cos(w) + 0.6*cos(k*w)*cos(2*w))
    }
    out <- integrate(gamma_k, lower = 0, upper = pi)$value
    return(out)
}

sapply(c(3:10), FUN = acvf)
```

Since the integrals here all equal zero for a sequence of $k > 2$, then $\gamma(k)$ = 0
for $k > 2$. $\medskip$ 

__Part e)__ $\medskip$

As such, we have the following autocovariance function

$$
\gamma(k) = \left\{
    \begin{array}{lr}
        1/8, & \text{for } k = 0\\
        2/25, & \text{for } k = 1 \\
        3/80, & \text{for } k = 2 \\
        0, & \text{for } k > 2 
    \end{array}
\right.
$$

where $k \in \mathbb{Z}$.

## Question 3

__Part a)__

```{r, echo = TRUE, message = FALSE}
pelts <- read_table("./data/pelt.txt", col_names = FALSE) %>%
    rename(sales = `X1`) %>%
    mutate(sales = sales / 1000,
           year = seq(year("1857-01-01"), year("1910-01-01")),
           t = year - 1856) %>%
    relocate(c(year, t), .before = sales)

pelts_chart <- ggplot(
    data = pelts,
    aes(x = year, y = sales)
) +
    geom_line(
        linewidth = 1.2,
        color = "#FF8833"
) +
    labs(
        x = "Year",
        y = "Sales",
        title = "Pelt Sales Between 1857 and 1910"
)

print(pelts_chart)
```

This looks like a pretty standard time series with an obvious seasonal effect.
However, the seasonal variation occurs over periods of years, rather than months
like we've usually seen so far. $\medskip$

__Part b)__

```{r, echo = TRUE, message = FALSE}
pelts_spec <- spec.pgram(pelts$sales, log = "no", plot = FALSE)

spectrum <- tibble(
    freq = seq(0, 0.5, length.out = length(pelts_spec$spec)),
    spec = pelts_spec$spec)

spectrum_chart <- ggplot(
    data = spectrum,
    aes(x = freq, y = spec)
) +
    geom_line(
        linewidth = 1.0,
        color = "#FF0652"
) +
    labs(
        x = "Frequency",
        y = "Spectrum",
        title = "Spectral Density of Pelt Sales"
)

print(spectrum_chart)
```

```{r, echo = TRUE, message = FALSE}
which.max(spectrum$spec)
```

In this case, $p = 6$ maximizes the spectrum, so the dominating frequency is

$$\omega_p = \frac{2\pi p}{N} = \frac{12\pi}{54}$$

__Part c)__

```{r, echo = TRUE, message = FALSE}
#' @param series A univariate time-series
#' @param p Fourier frequency constant, must be between 0 and N/2

fourier_freq <- function(series, p) {
    # Check inputs
    if (max(p) > length(series) / 2) {
        stop("p cannot be greater than N/2")
    }
    if (is.ts(series) == FALSE) {
        series <- as.ts(series)
    }
    # Generate frequency with given constant
    N <- length(series)
    w_p <- (2 * pi * p) / N

    # Return output when function is called
    return(w_p)
}

fourier_freq(series = pelts$sales, p = 8)
```

__Part d)__

```{r, echo = TRUE, message = FALSE}
significant_freqs <- function(w_p) {
    lm_fit <- lm(
        formula = log(sales) ~ cos(w_p * t) + sin(w_p * t),
        data = pelts
    )
    f_stat <- summary(lm_fit)$fstatistic
    p_val <- pf(
        q = f_stat[1], df1 = f_stat[2], df2 = f_stat[3],
        lower.tail = FALSE)
    return(p_val)
}

freqs <- fourier_freq(
    series = pelts$sales,
    p = seq(1, 27, length.out = 27)
)

sig_freqs <- tibble(
    p = seq(1, 27, length.out = 27),
    w_p = freqs,
    p_value = sapply(freqs, significant_freqs)) %>%
    filter(
        p_value < 0.05)

print(sig_freqs)
```

__Part e)__

```{r, echo = TRUE, message = FALSE}
lm_fit <- lm(
    log(sales) ~ cos(0.698 * t) + sin(0.698 * t),
    data = pelts)

coef(lm_fit)
```

__Part f)__

```{r, echo = TRUE, message = FALSE}
lm_data <- tibble(
    t = pelts$t,
    log_sales = log(pelts$sales),
    fitted_vals = fitted.values(lm_fit))

palette <- c("Sales" = "#FF0652", "Fourier Series" = "#000000")

lm_chart <- ggplot(
    data = lm_data,
    aes(x = t)
) +
    geom_line(
        aes(y = log_sales, color = "Sales"),
        linewidth = 1.2
) +
    geom_line(
        aes(y = fitted_vals, color = "Fourier Series"),
        linewidth = 1.2
) +
    labs(
        x = "t",
        y = "log(sales)",
        title = "Log-Sales and Fitted Fourier Series",
        color = "Series"
) +
    scale_x_continuous(
        n.breaks = 10
) +
    scale_color_manual(
        values = palette
)

print(lm_chart)
```