---
title: "Assignment 1"
author: "Matthew Kielar"
date: "`r Sys.Date()`"
output:
    bookdown::pdf_document2:
        toc: FALSE
        number_sections: false
---

## Question 1

__Part a)__

```{r load-packages, echo = FALSE, message = FALSE}
# Load in necessary libraries
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r plot-theme, include = FALSE, echo = FALSE}
plot_theme <- theme(
    plot.title = element_text(
        size = 16, face = "bold", hjust = 0,
        margin = margin(b = 0.25, unit = "cm")),
    plot.subtitle = element_text(
        size = 12, hjust = 0),
    axis.title.x = element_text(
        size = 12, face = "bold", angle = 0,
        margin = margin(r = 0.25, unit = "cm")),
    axis.title.y = element_text(
        size = 12, face = "bold", angle = 90,
        margin = margin(r = 0.25, unit = "cm")),
    axis.text = element_text(
        size = 12),
    axis.ticks.length = unit(0.25, "cm"),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_line(
        size = 0.1, color = "gray50"),
    panel.grid.major.y = element_line(
        size = 0.1, color = "gray50"),
    panel.background = element_rect(
        fill = "#FFFFFF"),
    panel.border = element_blank(),
    plot.background = element_rect(
        fill = "#FFFFFF"),
    legend.background = element_rect(
        fill = "#FFFFFF"),
    legend.key = element_rect(
        fill = "#FFFFFF"),
    legend.title = element_text(
        size = 12, face = "bold"),
    legend.text = element_text(
        size = 12),
    legend.justification = "left",
    legend.position = "top"
)

theme_set(plot_theme)
```

```{r, echo = TRUE, message = FALSE, fig.height = 4, fig.width = 8}
co2_mm_gl <- read_csv("./data/co2_mm_gl.csv", col_names = TRUE)

co2_chart <- ggplot(
    data = co2_mm_gl,
    aes(x = year, y = average)
) +
    geom_line(
        linewidth = 1.2,
        color = "#0F5499"
) +
    labs(
        x = "Year",
        y = "CO2 Concentration (ppm)",
        title = "Surface CO2 Concentration Since 1979"
)

print(co2_chart)
```

At first glance there is an obvious trend, as evidenced by the steady increase in CO2 levels
throughout the years. Seasonal variation is present too, as indicated by regular dips and spikes
at the end of each year, so in this case an additive model seems to be appropriate. As far as 
stationarity goes, it's clear that the series is not stationary, since the steady increase
seen in the chart almost surely means that the series does not have a constant mean. $\medskip$

__Part b)__ 

```{r, echo = TRUE, message = FALSE, fig.height = 5, fig.width = 8}
co2_train <- co2_mm_gl %>%
    filter(year < 2020) %>%
    select(average) %>%
    mutate(time = seq(1, length(average))) %>%
    relocate(time, .before = average)

co2_test <- co2_mm_gl %>%
    filter(year > 2019) %>%
    select(average) %>%
    mutate(time = seq(493, 492 + length(average))) %>%
    relocate(time, .before = average)

moving_avg <- decompose(
    ts(co2_train$average, start = 1979, frequency = 12),
    type = "additive")

loess <- stl(
    ts(co2_train$average, start = 1979, frequency = 12),
    s.window = "periodic")$time.series

plot(moving_avg, col = "#990F3D", lwd = 2)
plot(loess, col = "#990F3D", lwd = 2, main = "LOESS Decomposition")
```

__Part c)__ $\medskip$

```{r echo = TRUE, message = FALSE}
co2_train$moving_avg <- moving_avg$trend

co2_train <- co2_train %>%
    mutate(loess = as_tibble(loess)$trend)

lm_fit1 <- lm(formula = moving_avg ~ time, data = co2_train)
lm_fit2 <- lm(formula = loess ~ time, data = co2_train)
```

```{r, echo = TRUE}
lmtest::coeftest(lm_fit1)
```

\clearpage

```{r, echo = TRUE}
lmtest::coeftest(lm_fit2)
```

The fitted models for the two decomposition methods are as follows:

$$
\begin{aligned}
\text{MA}&: \text{CO2} = 332.9 + 0.1488t \\
\text{LOESS}&: \text{CO2} = 332.9 + 0.1491t
\end{aligned}
$$

The model specifications for both trends are extremely similar. The intercepts
are almost exactly the same, and $\hat{\beta_1}$ is only marginally different. Coefficients
for both models are significant at the 95% level, indicating that there is a positive 
linear relationship between trend value and time. $\medskip$

Based on this fact, it seems reasonable to want to make predictions given the statistical
significance of the trend component as well the fact that the weighted moving average via
LOESS shows a consistent increase over time.

__Part d)__ $\medskip$

```{r, echo = TRUE}
mean_sq_err <- function(obs, pred) {
    sq_diff <- (obs - pred)^2
    mse <- mean(sq_diff)
    return(mse)
}

moving_avg_preds <- predict.lm(lm_fit1, newdata = co2_test)
loess_preds <- predict.lm(lm_fit2, newdata = co2_test)

mean_sq_err(co2_test$average, moving_avg_preds)
mean_sq_err(co2_test$average, loess_preds)
```

Since LOESS smoothing has a slightly lower out-of-sample MSE, then it's likely the better
option if the priority is solely to make the most accurate predictions.

\clearpage

## Question 2

__Part a)__ $\medskip$

```{r, echo = TRUE, message = FALSE, fig.height = 5, fig.width = 8}
canada_gdp <- read_csv("./data/canadaGDP.csv", col_names = TRUE) %>%
    rename(Date = `...1`) %>%
    mutate(Date = ymd(Date))

gdp_chart <- ggplot(
    data = canada_gdp,
    aes(x = Date, y = GDP)
) +
    geom_line(
        linewidth = 1.2,
        color = "#0F5499"
) +
    labs(
        x = "Date",
        y = "GDP",
        title = "Canadian GDP (Monthly Adjusted) Since 1997"
) +
    scale_x_date(
        date_breaks = "3 years", date_labels = "%Y"
)

print(gdp_chart)
```

The series is likely non-stationary since it does not exhibit a constant mean across all
values of $t$. Given the constant increase in GDP each year, it seems reasonable to assume
that the mean is not constant.

__Part b)__

```{r, echo = TRUE}
sequential <- tibble(
    Date = canada_gdp$Date[2:276],
    Growth = diff(log(canada_gdp$GDP), lag = 1)
)

annual <- tibble(
    Date = canada_gdp$Date[13:276],
    Growth = diff(log(canada_gdp$GDP), lag = 12)
)

sequential_chart <- ggplot(
    data = sequential,
    aes(x = Date, y = Growth)
) +
    geom_line(
        linewidth = 1.2,
        color = "#990F3D"
) +
    labs(
        x = "Date",
        y = "GDP Growth Rate",
        title = "Canadian GDP Growth Rate (Sequential)"
)

annual_chart <- ggplot(
    data = annual,
    aes(x = Date, y = Growth)
) +
    geom_line(
        linewidth = 1.2,
        color = "#990F3D"
) +
    labs(
        x = "Date",
        y = "GDP Growth Rate",
        title = "Canadian GDP Growth Rate (Annual)"
)
```

```{r, echo = TRUE, fig.height = 4, fig.width = 8}
print(sequential_chart)
print(annual_chart)
```

The series for the sequential log-growth rate looks like it's stationary since the
mean/expected value appears to be roughly constant across all values of $t$. Annualized
log-growth on the other hand doesn't look stationary since there's a slight downward trend
at first glance, indicating that the mean might not be constant. $\medskip$

If further analysis is desired, then the sequential series is preferable due to stationarity
being more likely, but one drawback is that the excessive noise makes it harder for the chart
to convey information that could be of interest to another audience. $\medskip$

\clearpage

__Part c)__ 

```{r, echo = TRUE}
acf_sequential <- acf(
    x = ts(sequential$Growth),
    lag.max = nrow(sequential) - 1,
    main = "ACF of Sequential Series",
    lwd = 1.0)

acf_annual <- acf(
    x = ts(annual$Growth),
    lag.max = nrow(annual) - 1,
    main = "ACF of Annual Series",
    lwd = 1.0)
```

The correlogram for the sequential series is rather unremarkable. It oscillates between positive
and negative values that appear to asymptotically approach zero as the lag increases. For the 
annualized series, the ACF is clustered together at multiple intervals, which probably isn't a 
surprise since the annualized series displays far less random noise than the sequential series.

\clearpage

## Question 3

```{r, echo = TRUE, warning = FALSE, message = FALSE}
acf_simulation <- function(n = 1000, sd = 1, lag = 1, max_iter) {
    stopifnot(is.numeric(max_iter))

    # Pre-allocate space
    acf <- double(max_iter)

    # Run simulation
    set.seed(123)
    for (i in 1:max_iter) {
        z_t <- rnorm(n = n, mean = 0, sd = sd)
        acf[i] <- acf(z_t, plot = FALSE)$acf[lag + 1]
    }
    results <- tibble(iteration = 1:max_iter, acf)
    return(results)
}

simul <- acf_simulation(n = 1000, sd = 1, lag = 1, max_iter = 5000)

simul_chart <- ggplot(
    data = simul,
    aes(x = acf)
) +
    geom_histogram(
        aes(y = ..density..),
        color = "#000000",
        fill = "#FF7FAA"
) +
    stat_function(
        fun = dnorm,
        args = list(
            mean = -0.001, sd = sqrt(0.001)),
        linewidth = 1.2,
        color = "#000000"
) +
    labs(
        x = "Autocorrelation Coefficient (ACF)",
        y = "Density",
        title = "Distribution of ACF for 5000 WN Processes"
) +
    scale_x_continuous(
        limits = c(-0.12, 0.12)
)

print(simul_chart)
```

Based on the chart, the empirical density approxmiates the theoretical density really
well. Both are centered around the same mean and exhibit the same variance. If the simulation
were to be run for an even greater number of iterations, then the histogram would provide
an even better fit to the specified normal p.d.f. 